---
title: 'Lab 3: Visualization with ggplot2'
output: html_document
---

This week we introduced how to make more esthetically pleasing plots using `ggplot2`. For the remainder of the course you'll be expected to use `ggplot2` any time you create a plot. Many more plotting examples will be presented in class, but let's get more practice now.

Recall this figure from the `intro-to-ggplot.Rmd` lecture file:

```{r, message=FALSE, echo=FALSE}
library(ggplot2)
library(dplyr)
library(dslabs)
murders %>% ggplot() + 
            geom_point(aes(population/10^6, total), size = 3) +
            geom_text(aes(population/10^6, total, label = abb), nudge_x = 2)
```


Remake the plot above but 

1. flip the variables so that `total` is in the x-axis and `population` is in the y-axis 
2. show the state names instead of the abbreviation
3. make the points smaller

__Solution:__
```{r, message=FALSE}
data(murders)
murders %>% ggplot() + 
            geom_point(aes(x=total, y=population/10^6), size = 0.5) + 
            geom_text(aes(x=total, y=population/10^6, label = state))
```

### Grids of plots
Rather than using `facet_wrap` or `facet_grid`, there is another option for graphing plots next to each other: the `gridExtra` package permits us to do that. First create and name all of the plots you wish to show in the same window. Then use the function `grid.arrange()` and specificy which plots you want to include and how many columns you want to organize them in.

Below is an example using the `heights` data set. We define `p` to be a histogram of male heights and `p1`, `p2`, and `p3` to be the same histogram but with different binwidths i.e. different widths for the rectangles that make up the histogram.

```{r}
p <- heights %>% filter(sex == "Male") %>% 
                 ggplot(aes(x = height)) 

p1 <- p + geom_histogram(binwidth = 1, fill = "blue", col="black")
p2 <- p + geom_histogram(binwidth = 2, fill = "blue", col="black")
p3 <- p + geom_histogram(binwidth = 3, fill = "blue", col="black")
```

To print them all side-by-side, we can use the function `grid.arrange` in the `gridExtra` package:

```{r, warning=FALSE, message=FALSE}
library(gridExtra)
grid.arrange(p1,p2,p3, ncol = 3)
```

Here, `ncol` specifies how many columns I want to arrange my plot in.


4. Using the `gapminder` dataset, filter the data frame so only data from the year 2015 is kept. 

__Solution__
```{r}
data(gapminder)
gapminder <- gapminder %>% filter(year == "2015")
```


5. Create a histogram of `fertility`, and a second histogram of `life expectancy`. Then use `grid.arrange()` to plot the histograms on top of each other in the same window.

__Solution__
```{r, message=FALSE, warning=FALSE}
p1 <- gapminder %>% ggplot(aes(x = fertility)) +          
                    geom_histogram(color = "black")

p2 <- gapminder %>% ggplot(aes(x = life_expectancy))  +                            
                    geom_histogram(color = "black")

grid.arrange(p1,p2, ncol = 1)
```


## Vaccines

For homework #2 you'll be working with the `us_contagious_diseases` dataset in the `dslabs` package. The data were collected, organized and distributed by the [Tycho Project](https://www.tycho.pitt.edu/). They include weekly reported counts data for seven diseases from 1928 to 2011, from all fifty states. We include the yearly totals in the `dslabs` package:
```{r}
library(dslabs)
data("us_contagious_diseases")
```

6. Use the `us_contagious_disease` dataset and `dplyr` tools to create an object called `dat` that stores only the Smallpox data, includes a per 10,000 people rate, and removes Alaska and Hawaii since they only became states in the late 1950s. Do not include years in which cases were not reported in 10 or more weeks.

__Solution__
```{r}
dat <- us_contagious_diseases %>% 
       filter(!state %in% c("Hawaii", "Alaska") & 
       disease == "Smallpox" & weeks_reporting >= 10) %>%
       mutate(rate = count / population * 10000)
```


7. Calculate the US average number of cases of Smallpox per 10,000 people for each year. Do not include years in which cases were not reported.

__Solution__
```{r}
avg <- us_contagious_diseases %>%
       filter(disease == "Smallpox") %>% 
       group_by(year) %>%
       summarize(us_rate = sum(count, na.rm=TRUE)/sum(population, na.rm=TRUE)*10000)
```


8. Make a the time series plot of the number of cases of Smallpox per 10,000 people for all states and add a line for the US average.

__Solution__
```{r}
dat %>% ggplot() + geom_line(aes(year, rate, group = state),  
                             color = "grey50", 
                             show.legend = FALSE, 
                             alpha = 0.2, size = 1) +
                   geom_line(mapping = aes(year, us_rate),
                             data = avg, 
                             size = 1, 
                             color = "black") +
                   scale_y_continuous(trans = "sqrt") +
                   ggtitle("Cases per 10,000 by state") + 
                   xlab("") + 
                   ylab("")
```

9. For the state of California, make a time series plot showing rates for all diseases. Include only years with 10 or more weeks reporting. Use a different color for each disease.

__Solution__
    
```{r}
us_contagious_diseases %>% filter(state == "California" & weeks_reporting >= 10) %>%
                           mutate(rate = count/population*10000) %>%
                           ggplot(aes(year, rate, color = disease)) + 
                           geom_line()
```
