---
title: 'Lab 5: Shiny'
output: html_document
---

This lab will guide through building a Shiny app using the `murders` dataset from the `dslabs` package.

## Step 1

Create a basic Shiny app that includes a `selectInput` input widget that allows the user to choose one of the 4 regions (Northeast, North Central, West, or South) and outputs a scatterplot with `total` murder on the y-axis and `population` on the x-axis. Don't forget to load the `shiny`, `dslabs`, `dplyr` and `ggplot2` libraries as well as the `murders` dataset.

* Solution: 

```{r, eval=FALSE}
library(shiny)
library(dslabs)
library(dplyr)
library(ggplot2)
data("murders")

# Define ui
ui <- fluidPage( # Dropdown menu that allows the user to choose a region
                 selectInput(inputId = "region", label = "Select a region",
                             choices = as.list(levels(murders$region))),
                 plotOutput("scatterPlot")
)

# Define server
server <- function(input, output){
                   output$scatterPlot <- renderPlot({
                   		murders %>% filter(region == input$region) %>%
                   								ggplot(aes(x = population, y = total)) +
                  							  geom_point() +
                   								xlab("Total number of murders") +
                   								ylab("Population")
    })
}

shinyApp(ui = ui, server = server)

```


## Step 2
Points in the lower left of the plot are bunched up for most of the regions. To make it easier to view these points, transform the y-axis using the `log10` transform.

* Solution: 

Only the `server` function is changed. Advanced: `limits` and `breaks` were used to keep the axis the same for each region's plot.

```{r, eval=FALSE}
server <- function(input, output){
    output$scatterPlot <- renderPlot({
        murders %>% filter(region == input$region) %>%
            ggplot(aes(x = population/10^6, y = total)) +
            geom_point() +
            scale_x_log10(limits = c(1, 30), breaks = c(1, 3, 10, 30)) +
            scale_y_log10(limits = c(1, 1000), breaks = c(10, 100, 300, 500, 1000)) +
            xlab("Population in Millions(log scale)") +
            ylab("Total Number of Murders (log scale)")
    })
}
```

## Step 3

We have a working app but we haven't tried out different layouts. Using the `siderbarLayout`, `sidebarPanel` and `mainPanel` functions to have the input appear on the left and the scatter plot on the right. Include a title for your app using the `titlePanel` function.

* Solution:

Only the `ui` function changed.

```{r, eval=FALSE}
ui <- fluidPage( 
    titlePanel("Data Exploration with Murder Data"),
    
    sidebarLayout(
        sidebarPanel(
                # Dropdown menu that allows the user to choose a region
                selectInput(inputId = "region", label = "Select a region",
                            choices = as.list(levels(murders$region)))
                ),
        
        mainPanel(
                plotOutput("scatterPlot")
                )
    )
)
```


## Step 4
Now include an input that allows the user to decide what color they would like the points to be. Include the following colors: `black`, `blue`, `red` and `green`.

* Solution:

Both the `ui` and `server` functions change.

```{r, eval=FALSE}
ui <- fluidPage( 
    titlePanel("Data Exploration with Murder Data"),
    
    sidebarLayout(
        sidebarPanel(
            # Dropdown menu that allows the user to choose a region
            selectInput(inputId = "region", label = "Select a region",
                        choices = as.list(levels(murders$region))),
            
            selectInput(inputId = "color", label = "Choose a color",
                        choices = c("black", "blue", "red", "green")
                        )
            ),
        
        mainPanel(
            plotOutput("scatterPlot")
        )
    )
)

# Define server
server <- function(input, output){
    output$scatterPlot <- renderPlot({
        murders %>% filter(region == input$region) %>%
            ggplot(aes(x = population/10^6, y = total)) +
            geom_point(color = input$color) +
            scale_x_log10(limits = c(1, 30), breaks = c(1, 3, 10, 30)) +
            scale_y_log10(limits = c(1, 1000), breaks = c(10, 100, 300, 500, 1000)) +
            xlab("Population in Millions(log scale)") +
            ylab("Total Number of Murders (log scale)")
    })
}
```

## Step 5
I want to give the user the option to view the scatter plot we already have or a boxplot of total number of murders for each region. Add an input that would allow this.

* Solution:

Both the `ui` and `server` function change.
```{r, eval=FALSE}
ui <- fluidPage( 
    titlePanel("Data Exploration with Murder Data"),
    
    sidebarLayout(
        sidebarPanel(
            # Dropdown menu that allows the user to choose a region
            selectInput(inputId = "region", label = "Select a region",
                        choices = as.list(levels(murders$region))),
            
            # Dropdown menu that allows user to choose a color
            selectInput(inputId = "color", label = "Choose a color",
                        choices = c("black", "blue", "red", "green")
            ),
            
            # Radio buttons for the user to choose a scatter plot or boxplot
            radioButtons(inputId = "type", label = "What kind of plot?",
                         choices = c("Scatter plot", "Boxplot"))
        ),
        
        mainPanel(
            # I changed the name of the output since it can be 2 different things
            plotOutput("Plot") 
        )
    )
)

# Define server
server <- function(input, output){
        output$Plot <- renderPlot({
            if (input$type == "Scatter plot"){
                murders %>% filter(region == input$region) %>%
                    ggplot(aes(x = population/10^6, y = total)) +
                    geom_point(color = input$color) +
                    scale_x_log10(limits = c(1, 30), breaks = c(1, 3, 10, 30)) +
                    scale_y_log10(limits = c(1, 1000), breaks = c(10, 100, 300, 500, 1000)) +
                    xlab("Population in Millions(log scale)") +
                    ylab("Total Number of Murders (log scale)")
            } else{
                murders %>% filter(region == input$region) %>%
                    ggplot(aes(x = region, y = total)) +
                    geom_boxplot(color = input$color) +
                    xlab("") +
                    ylab("Number of murders")
            }
        })
    
}
```


## Step 6
Suppose I also want to know the mean number of murders for a particular region. Add an output that shows the mean number of murders for a chosen region using the `verbatimTextOutput()` function in the `ui` function. Use the `renderText({ })` function in the `server` function to have the mean ouput in the app.

* Solution:

Both the `ui` and `server` function change.
```{r, eval=FALSE}
ui <- fluidPage( 
    titlePanel("Data Exploration with Murder Data"),
    
    sidebarLayout(
        sidebarPanel(
            # Dropdown menu that allows the user to choose a region
            selectInput(inputId = "region", label = "Select a region",
                        choices = as.list(levels(murders$region))),
            
            # Dropdown menu that allows user to choose a color
            selectInput(inputId = "color", label = "Choose a color",
                        choices = c("black", "blue", "red", "green")),
            
            # Radio buttons for the user to choose a scatter plot or boxplot
            radioButtons(inputId = "type", label = "What kind of plot?",
                         choices = c("Scatter plot", "Boxplot"))
            ),
        
        mainPanel(
            # I changed the name of the output since it can be 2 different things
            plotOutput("Plot"),
            verbatimTextOutput("mean")
        )
    )
)

# Define server
server <- function(input, output){
    output$Plot <- renderPlot({
        if (input$type == "Scatter plot"){
            murders %>% filter(region == input$region) %>%
                ggplot(aes(x = population/10^6, y = total)) +
                geom_point(color = input$color) +
                scale_x_log10(limits = c(1, 30), breaks = c(1, 3, 10, 30)) +
                scale_y_log10(limits = c(1, 1000), breaks = c(10, 100, 300, 500, 1000)) +
                xlab("Population in Millions(log scale)") +
                ylab("Total Number of Murders (log scale)")
        } else{
            murders %>% filter(region == input$region) %>%
                ggplot(aes(x = region, y = total)) +
                geom_boxplot(color = input$color) +
                xlab("") +
                ylab("Number of murders")
        }
    })
    
    output$mean <- renderText({
        avg <- murders %>% filter(region == input$region) %>%
                summarize(avg = mean(total)) %>% .$avg
            
        paste0("Mean number of murders in the ", input$region, " region: ", avg)
    })
    
}
```


## Step 7
We use the same subset of the `murders` data set in 3 different outputs in our app and redefine the subset each time the input changes. This is inefficient and takes up more lines of code. Make the subset of the `murders` data frame a reactive expression. Then use that expression in each of the 3 outputs.

* Solution:

Only the `server` function changed.

```{r, eval=FALSE}
server <- function(input, output){
    dat <- reactive(murders %>% filter(region == input$region))
    
    output$Plot <- renderPlot({
        if (input$type == "Scatter plot"){
            dat() %>%
                ggplot(aes(x = population/10^6, y = total)) +
                geom_point(color = input$color) +
                scale_x_log10(limits = c(1, 30), breaks = c(1, 3, 10, 30)) +
                scale_y_log10(limits = c(1, 1000), breaks = c(10, 100, 300, 500, 1000)) +
                xlab("Population in Millions(log scale)") +
                ylab("Total Number of Murders (log scale)")
        } else{
            dat() %>%
                ggplot(aes(x = region, y = total)) +
                geom_boxplot(color = input$color) +
                xlab("") +
                ylab("Number of murders")
        }
    })
    
    output$mean <- renderText({
        avg <- dat() %>%
            summarize(avg = mean(total)) %>% .$avg
        
        paste0("Mean number of murders in the ", input$region, " region: ", avg)
    })
    
}
```



## Step 8 (Optional)
Add some text before the input widgets that gives some background/context to your app. Use the paragraph function `p()`. You may want to also use the `br()` function to add some space between the text and input widgets.

* Solution:

Only the `ui` function changes.

```{r, eval=FALSE}
ui <- fluidPage( 
    titlePanel("Data Exploration with Murder Data"),
    
    sidebarLayout(
        sidebarPanel(
            p("The US is a large and diverse country with 50 very different states as well as the 
              District of Columbia (DC). Here we use the", code("murders"), "dataset to gain some insight
              into the variability in number of murders by region."),
            
            br(),
            
            # Dropdown menu that allows the user to choose a region
            selectInput(inputId = "region", label = "Select a region",
                        choices = as.list(levels(murders$region))),
            
            # Dropdown menu that allows user to choose a color
            selectInput(inputId = "color", label = "Choose a color",
                        choices = c("black", "blue", "red", "green")),
            
            # Radio buttons for the user to choose a scatter plot or boxplot
            radioButtons(inputId = "type", label = "What kind of plot?",
                         choices = c("Scatter plot", "Boxplot"))
        ),
        
        mainPanel(
            # I changed the name of the output since it can be 2 different things
            plotOutput("Plot"),
            verbatimTextOutput("mean")
        )
    )
)
```


## Step 9 (Optional)
Create the same app but with a different layout using the `multiRow` and `column` functions.

* Solution:

Here is one example. Only the `ui` function changes.

```{r, eval=FALSE}
ui <- fluidPage( 
    titlePanel("Data Exploration with Murder Data"),
    br(), # add some space 
    # First row
    fluidRow(
        column(3,
              p("The US is a large and diverse country with 50 very different states as well as the 
              District of Columbia (DC). Here we use the", code("murders"), "dataset to gain some insight
              into the variability in number of murders by region.")
        ),
        column(3, selectInput(inputId = "region", label = "Select a region",
                              choices = as.list(levels(murders$region)))
        ),
        column(3, selectInput(inputId = "color", label = "Choose a color",
                              choices = c("black", "blue", "red", "green"))
        
        ),
        column(3, radioButtons(inputId = "type", label = "What kind of plot?",
                               choices = c("Scatter plot", "Boxplot"))
        )
    ),
    br(), # add some space 
    # Second row
    fluidRow(
        column(6, plotOutput("Plot")),
        column(6, verbatimTextOutput("mean"))
    )
)
```


