---
title: 'Lab 4: More Visualization with ggplot2'
output: html_document
---

```{r libraries}
library(dslabs)
library(dplyr)
library(ggplot2)
```

In this lab, we will be working more with the vaccination data and we
will be practicing using ggplot2. The examples here attempt to show how
a data scientist would approach a new dataset, and are intended to
illustrate how `ggplot2` can be flexibly manipulated to create a variety
of visualizations.

# Problem 1: Weeks reporting

In this problem, we will explore the `weeks_reporting` column in the
vaccines data and its impact on the rate calculation.

```{r load_diseases}
data(us_contagious_diseases)
```
a. Look at the help documentation for `us_contagious_diseases` and
describe what `weeks_reporting` represents.

*Solution*: The `weeks_reporting` column counts the number of weeks
counts were reported that year.

b. Draw a histogram of the `weeks_reporting` column for all states, all
years, and all diseases combined.

```{r hist_weeks}
# your code here
```

c. The histogram above aggregates the `weeks_reporting` for all states,
all years, and all diseases combined. It could be possible that there
are some diseases where the number of weeks reporting could be close to
52, and some diseases where the number of weeks reporting could be much
less. Does the number of weeks reporting vary by disease? Use the
`facet_wrap` function to explore this.


```{r hist_disease}
# your code here
```

d. Because the histogram above aggregates data on all states, it's
possible there are differences in weeks reporting by states as well. To
answer this question, we can draw a boxplot by plotting the weeks
reporting by state, and facetting by disease.

```{r boxplot, fig.width = 8, fig.height = 8}
# your code here
```

e. If the above plot is hard to read, we might want to focus on one
disease in particular. How can we revise the code in part (d) to only
plot Pertussis?

```{r boxplot_pertussis}
# your code here
```

# Problem 2: Improved Rate Calculation

Recall that last time we calculated the rate for each disease in each
state in each year as the count divided by the population. Now that
we've identified that we need to take into account differential weeks
reporting in calculating the rate, let us amend the rate calculation.

a. Subset the data to exclude Alaska and Hawaii and include states-years
with weeks reporting greater than 0.

```{r filter}
# your code here
```
b. Calculate the rate per 100,000, not taking into account weeks
reporting. Call this the unadjusted rate.

```{r raw_rate}
# your code here
```

b. Now, take into account weeks reporting for vaccination data to
calculate the rate. Call this the adjusted rate.

```{r adjust_rate}
# your code here
```

c. Compare the adjusted vs. unadjusted rate to see if there is a
difference. Make a scatter plot with one quantity on the x-axis and one
quantity on the y-axis. Color each point by the number of weeks
reporting. Include the identity line so we can more easily make
comparisons.

In statistics, we often want to compare two quantities to see how
they're related. Do we see a relationship here? Looking back at the
derivation of the new rate, is it clear why the new rate is greater than
or equal to the previously calculated rate? Is there a relationship
between the difference and the number of weeks reporting?

```{r compare}
# your code here
```

# Problem 3

Now, we can apply the same principle to calculate the overall rate for
each disease for the entire US.

a. For each disease and year, compute the overall rate for the entire
US. Hint: You may want to use the `group_by` function.

```{r us_rate}
# your code here
```

b. Draw a plot of the correctly calculated overall rate for the entire
US for each disease.

```{r us_plot}
# your code here
```

# Problem 4

In this problem, we will explore some other types of visualization and
data manipulation using the gapminder data.

a. Create a new dataframe called `gapminder_africa` which only includes
countries from the gapminder dataset that are in the continent of
Africa.

```{r gapminder_africa}
data(gapminder)
# your code here
```

b. We will explore the changes in population for each country over time.
First, create a line graph for each country's population over time.

```{r gapminder_pop}
# your code here
```

c. We can see that the distribution of populations is highly
skewed. There are a bunch of countries with lower populations
that are all bunched up. We can fix this by performing a log10
transformation of the y-axis.

```{r gapminder_log10}
# your code here
```

d. In the line graph in part (c), it is hard to see individual
countries. Instead, we can make a tile plot of the population. Check out
the `geom_tile` command and produce a tile plot of each country's
population over time. We can color each tile using the `fill` aesthetic.

```{r gapminder_tile}
# your code here
```

e. `ggplot2`'s default color scheme looks a little ugly. We can change
that! There is a built-in colorscheme called viridis that we can try.
Look up `scale_fill_viridis_c` in the help. We can use `viridis_c` for
continuous data and `viridis_d` for discrete data. Here, because we are
using `fill` to color each tile, we need to use `scale_fill_viridis_c`.
Note, we can also use the approach from lecture to color this plot.

```{r gapminder_tile_color}
# your code here
```

Also note that we can use the `name` parameter in the `scale_` functions
to set the title of the legend.

f. The previous plot was hard to read because the countries were in
alphabetical order. Instead, maybe we can order the countries by the
maximum population (i.e. the population they reached in the most recent
year available in gapminder).

First, use `group_by` and `summarize` to calculate the maximum
population (over years) for each country.

```{r gapminder_summary}
# your code here
```

Note, because of missing data, we need to use `max` with `na.rm = TRUE`
to remove the NAs.

g. Now, using the `reorder` function, create a new column
`country_reorder` in the `gapminder_africa` data that reorders the
countries by their maximum population over all the years in the
gapminder data.

```{r gapminder_order}
# your code here
```

h. Finally, use the `country_reorder` variable to make a revised version
of the plot in part (e).

Also, note that `ggplot2` `scale` functions also take a `trans`
argument, so you can apply a transformation to the scale. Apply a log10
transformation to the fill.

```{r gampinder_tile_reorder}
# your code here
```
